---
title: "Week 7: Spatial Linear Models"
author: "Helene Wagner and Yessica Rico"
date: "`r Sys.Date()`"
show_toc: true
output:
  knitr:::html_vignette:
    toc: yes
    fig_width: 4 
    fig_height: 3.5
vignette: >
  %\VignetteIndexEntry{Week 7: Spatial Linear Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
## 1. Overview of Worked Example

### a) Goals 

This worked example shows:

- How to test regression residuals for spatial autocorrelation.
- How to fit a model with spatially autocorrelated errors (GLS).
- How to fit a spatial simultaneous autoregressive error model (SAR).
- How to perform spatial filtering with Moran eigenvector maps (MEM).

### b) Data set

Here we analyze population-level data of the wildflower *Dianthus carthusianorum* (common name: Carthusian pink) in 65 calcareous grassland patches in the Franconian Jura, Germany (Rico et al. 2013): 

- **Dianthus**: Spatial point data frame with population-level data (patch characteristics, grazing regime, genetic diversity, 15 alternative connectivity indices Si), included in package 'LandGenCourse'. To load the data, type (without quotes): 'data(Dianthus)'. For a definition of the variables, type: '?Dianthus'.

### c) Required R libraries

All required packages should have been installed already when you installed 'LandGenCourse'.

Note: the function 'library' will always load the package, even if it is already loaded, whereas 'require' will only load it if it is not yet loaded. Either will work.

```{r message=FALSE, warning=TRUE}
require(LandGenCourse)
#require(here)
#require(spdep)
#require(nlme)
#require(lattice)
#require(MuMIn)
require(ggplot2)
source(system.file("extdata", "panel.cor.r", 
                            package = "LandGenCourse"))
```

Package 'spmoran' not automatically installed with 'LandGenCourse':

```{r message=FALSE, warning=TRUE}
if(!require(spmoran)) install.packages("spmoran")
#require(spmoran)
```

## 2. Explore data set

We will model allelic richness 'A' as a function of the following predictors:

- **IBD**: connectivity index Si ('Eu_pj') based on Euclidean distance between source and focal patch. This represents a hypothesis of isolation by distance (IBD).
- **IBR**: connectivity index Si ('Sheint_pj') based on the number of continuously or intermittently grazed patches between source and focal patch. This represents a hypothesis of isolation by resistance (IBR). Specifically, this model assumes connectivity via sheep-mediated seed dispersal, where seeds are likely to be transported from patch to patch within the same grazing system (shepherding route). Seeds are assumed to disperse most likely to the next patch (in either direction) along the grazing route, and less likely to more remote patches along the route.
- **PatchSize**: Logarithm of calcareous grassland patch size in ha. 

The connectivity indices Si were calculated for each focal patch i, integrating over all other patches j where the species was present (potential source patches) using Hanski's incidence function.

### a) Import data

```{r}
data(Dianthus)
```

Allelic richness 'A' was not calculate for populations with < 5 individuals. Here we extract only the patches with 'A' values, and the variables needed, and store them in a data frame 'Dianthus.df'.

### b) Create a map

Note: the coordinates 'x' and 'y' in Dianthus@coords are Gauss-Krueger projection GK4 as defined in Dianthus@proj4string. In addition, 'Longitude' and 'Latitude' are available as variables in Dianthus@data. This is useful for functions like 'qmplot' that expect latlon coordinates. 

```{r}
ggmap::qmplot(x =  Longitude, y = Latitude, data = Dianthus@data,
              source = "google", maptype = "terrain", zoom = 12)
```

As you can see from the map, most sites lie on the steep slopes between an upper and a lower Jurassic plateau. A few sites, south of road "S12216", lie at the forest edge on the upper plateau, typically in areas where the soil is too shallow to allow crop farming. With in the study area, all known sites were sampled. Additional sites are expected to be found mainly in the valley system in the Southwest. 

### c) Explore correlations

When fitting linear models, it is always a good idea to look at the correlations first. 

```{r fig.height=5, fig.width=7}
Dianthus.df <- data.frame(A=Dianthus@data$A, IBD=Dianthus@data$Eu_pj, 
                          IBR=Dianthus@data$Sheint_pj,
                          PatchSize=log(Dianthus@data$Ha), 
                          x=Dianthus@coords[,1], y=Dianthus@coords[,2])
Dianthus.df <- Dianthus.df[!is.na(Dianthus.df$A),]
dim(Dianthus.df)
pairs(Dianthus.df, lower.panel=panel.smooth, upper.panel=panel.cor,
      diag.panel=panel.hist)
```

Questions:

- How strong is the association between 'Eu_pj' and 'A'? What does this suggest about the hypothesis of IBD?
- How strong is the association between 'Sheint_pj' and 'A'? What does this suggest about the hypothesis of sheep-mediated gene flow (IBR)?
- Which variable seems to be a better predictor of allelic richness: patch size 'Ha' or the logarithm of patch size, 'PatchSize'?
- Is logHa correlated with 'IBD' or 'IBR'?
- Are any of the variables correlated with the spatial coordinates x and y?

Let's also check the association between patch size and population size:

```{r}
boxplot(log(Dianthus$Ha) ~ Dianthus$pop09, ylab="PatchSize (log(Ha))",
        xlab="Population size category")
```

Even though the population size categories were very broad, there appears to be a strong relationship between populations size (category) and (the logarithm of) patch size. 

Despite this relationship, connectivity models Si that only considered *Dianthus carthusianorum* presence/absence ('pj') in source patches 'j' were better supported than those Si models that took into account source patch area ('Aj') or population size ('Nj'). 

We can check this by calculating the correlation of allelelic richness 'A' with each of the 15 connectivity models 'Si' in the data set. 

```{r}
round(matrix(cor(Dianthus@data$A, Dianthus@data[,15:29], 
                 use="pairwise.complete.obs"), 5, 3, byrow=TRUE, 
           dimnames=list(c("Eu", "Shecte", "Sheint", "Shenu", "Forest"), 
                         c("pj", "Aj", "Nj"))),3)
```

- Correlations with 'A' are highest for the two 'IBR' models that assume seed dispersal over a limited number of patches along shepherding routes ('Shecte' and 'Sheint'. These two models include only continuously grazed, or both continuously and intermittently grazed patches, respectively. 
- Correlations for models that take in to account population size ('Nj') are only slightly lower, whereas those that use patch size ('Aj') as a proxy for the size of the seed emigrant pool had lower correlations. 

## 3. Test regression residuals for spatial autocorrelation

### a) Fit regression models

Here we fit three multiple regression models to explain variation in allelic richness:

- **mod.lm.IBD**: IBD model of connectivity 'Eu_pj'.
- **mod.lm.IBR**: IBR model shepherding connectivity 'Sheint_pj'.
- **mod.lm.PatchSize**: log patch size and IBR model.

```{r}
mod.lm.IBD <- lm(A ~ IBD, data = Dianthus.df)
summary(mod.lm.IBD)
```

This model does not fit the data at all!

```{r}
mod.lm.IBR <- lm(A ~ IBR, data = Dianthus.df)
summary(mod.lm.IBR)
```

This model fits much better. Let's check the residuals plots.

```{r fig.height=5.5, fig.width=7}
par(mfrow=c(2,2), mar=c(4,4,2,1))
plot(mod.lm.IBR)
par(mfrow=c(1,1))
```

The residuals show some deviation from a normal distribution. Specifically, the lowest values are lower than expected.

```{r}
mod.lm.PatchSize <- lm(A ~ PatchSize + IBR, data = Dianthus.df)
summary(mod.lm.PatchSize)
```

This combinde model explains more variation in allelic richness than the IBR model alone. Moreover, after adding PatchSizes, the IBR term is no longer statistically significant!

Has the distribution of residuals improved as well?

```{r fig.height=5.5, fig.width=7}
par(mfrow=c(2,2), mar=c(4,4,2,1))
plot(mod.lm.PatchSize)
par(mfrow=c(1,1))
```

Not really!

### b) Test for spatial autocorrelation (Moran's I):

Before we interpret the models, let's check whether the assumption of independent residuals is violated by spatial autocorrelation in the residuals. 

To calculate and test Moran's I, we first need to define neighbours and spatial weights. Here we use a Gabriel graph to define neighbours. 

We define weights in three ways (see Week 5 video and tutorial for code explanation): 

- **listw.gab**: 1 = neighbour, 0 = not a neighbour.
- **listw.d1**: inverse distance weights: neighbour j with weight 1/dij
- **listw.d2**: inverse squared distance weights: neighbour j with weight 1/dij^2

In each case, we row-standardize the weights with the option 'style = "W"'.

```{r}
xy <- data.matrix(Dianthus.df[,c("x", "y")])
nb.gab <- spdep::graph2nb(spdep::gabrielneigh(xy), sym=TRUE)
par(mar=c(0,0,0,0))
plot(nb.gab, xy)
listw.gab <- spdep::nb2listw(nb.gab)

dlist <- spdep::nbdists(nb.gab, xy)
dlist <- lapply(dlist, function(x) 1/x)
listw.d1 <- spdep::nb2listw(nb.gab, style = "W", glist=dlist)
dlist <- lapply(dlist, function(x) 1/x^2)
listw.d2 <- spdep::nb2listw(nb.gab, style = "W", glist=dlist)
```

Now we can quantify and test Moran's I for each variable to test for spatial autocorrelation in response and predictor variables. For now, we'll take the simple weights 'listw.gab'.

```{r}
spdep::moran.test(Dianthus.df$A, listw.gab)             
```
```{r}
spdep::moran.test(Dianthus.df$IBD, listw.gab)
```
```{r}
spdep::moran.test(Dianthus.df$IBR, listw.gab) 
```
```{r}
spdep::moran.test(Dianthus.df$PatchSize, listw.gab) 
```
Questions:

- Which variables showed statistically signficant spatial autocorrelation?
- Which variables showed the strongest autocorrelation? Is this surprising?

Next, let's test each model for autocorrelation in the residuals:

```{r}
spdep::lm.morantest(mod.lm.IBD, listw.gab) 
```
```{r}
spdep::lm.morantest(mod.lm.IBR, listw.gab)          
```
```{r}
spdep::lm.morantest(mod.lm.PatchSize, listw.gab)       
```

Quite a bit of the spatial autocorrelation in allelic richness can be explained by the spatial structure in the predictors IBR and PatchSize. There is still statistically significant spatial autocorrelation in the residuals, though it is not strong any more. 

## 4. Fit models with spatially correlated error (GLS) with package 'nlme'

One way to account for spatial autocorrelation in the residuals is to fit a Generalized Least Squares model (GLS) with a spatially autocorrelated error structure. 
See also: http://rfunctions.blogspot.ca/2017/06/how-to-identify-and-remove-spatial.html

The error structure in a GLS is defined in a geostatistical framework, based on a variogram and as a function of distance between observations. Hence we start with plotting an empirical variogram of the residuals, with a smooth line. Here we specify 'resType = "normalized", which means that the variogram will be fitted to the normalized residuals of the model. The expected value of the semivariance will be 1.

```{r}
mod.gls.PatchSize <- nlme::gls(A ~ IBR + PatchSize, data = Dianthus.df)
summary(mod.gls.PatchSize)


semivario <- nlme::Variogram(mod.gls.PatchSize, form = ~x  + y, 
                             resType = "normalized")
plot(semivario, smooth = TRUE)
lattice::trellis.focus("panel", 1, 1)
lattice::panel.abline(h=1)
lattice::trellis.unfocus() 
```

Question:

- What do you conclude from this empirical variogram? 
- Estimate the range of the variogram from the intersection of the smooth line with the horizontal line.
- Estimate the nugget effect from the intercept at Distance = 0.

We can ask R to fit different types of variogram models to this empirical variogram. The model family (e.g., exponential, gaussian, spherical) determines the general shape of the curve that will be fitted. With 'nugget=T', we indicate that a nugget effect should be fitted. 

```{r}
exponential.autocor <- nlme::gls( A ~ PatchSize + IBR, data = Dianthus.df, 
                            correlation = nlme::corExp(form = ~ x + y, nugget=T))

gaussian.autocor <- nlme::gls( A ~ PatchSize + IBR, data = Dianthus.df, 
                            correlation = nlme::corGaus(form = ~ x + y, nugget=T))

spherical.autocor <- nlme::gls( A ~ PatchSize + IBR, data = Dianthus.df, 
                            correlation = nlme::corSpher(form = ~ x + y, nugget=T))

#linear.autocor <- nlme::gls( A ~ PatchSize + IBR, data = Dianthus.df, 
#                            correlation = nlme::corLin(form = ~ x + y, nugget=T))

ratio.autocor <- nlme::gls( A ~ PatchSize + IBR, data = Dianthus.df, 
                            correlation = nlme::corRatio(form = ~ x + y, nugget=T))
```

Now we compare all models for which we did not get an error message:

```{r}
MuMIn::model.sel(mod.gls.PatchSize, exponential.autocor, gaussian.autocor, 
          spherical.autocor, ratio.autocor)     
```

The list sorts the models, with the best model on top. The last column 'weight' contains the model weight, which indicate how much support there is for each model, given all other models in the set (see Week 12). Here, the exponential model fitted best, though the ratio model and the model without a spatially correlated error structure fitted the data almost equally well.

```{r}
summary(exponential.autocor)  
```

The fitted model with the exponential error structure shows a significant effect for PatchSize but not for the IBR term.

We don't get an R-squared value directly, but we can calculate a pseudo R-squared from a regression of the response 'A' on the fitted values.

```{r}
summary(lm(A ~ fitted(exponential.autocor), data = Dianthus.df))$r.squared
summary(mod.lm.PatchSize)$r.squared
```

The pseudo R-squared is almost identical to the R-squared of the non-spatial 'lm' model. 

Let's check some residual plots - in this case, we have to construct them ourselves.

```{r fig.show='hold'}
plot(fitted(exponential.autocor), residuals(exponential.autocor))
abline(h=0,lty=3)

qqnorm(residuals(exponential.autocor))
qqline(residuals(exponential.autocor))
```

The normal probability plot still looks about the same.

```{r}
semivario <- nlme::Variogram(exponential.autocor, form = ~ x + y, 
                             resType = "normalized")
plot(semivario, smooth = TRUE)
lattice::trellis.focus("panel", 1, 1)
lattice::panel.abline(h=1)
lattice::trellis.unfocus() 
```

The variogam does look better!


## 5. Fit spatial simultaneous autoregressive error models (SAR) with package 'spdep'

An alternative way to account for spatial autocorrelation in the residuals is spatial regression with a simultaneous autoregressive error model (SAR).

The method 'errorsarlm' fits a simultaneous autoregressive model ('sar') to the error ('error') term of a 'lm' model. 

This approach is based on spatial neighbours and weights. We have already defined them in three versions of a 'listw' object. Let's see which one fits the data best.

```{r}
mod.sar.IBR.gab <- spdep::errorsarlm(A ~ PatchSize + IBR, data = Dianthus.df, 
                                 listw = listw.gab)
mod.sar.IBR.d1 <- spdep::errorsarlm(A ~ PatchSize + IBR, data = Dianthus.df, 
                                 listw = listw.d1)
mod.sar.IBR.d2 <- spdep::errorsarlm(A ~ PatchSize + IBR, data = Dianthus.df, 
                                 listw = listw.d2)

MuMIn::model.sel(mod.lm.IBR, mod.sar.IBR.gab, mod.sar.IBR.d1, mod.sar.IBR.d2) 
```

The best model ('mod.sar.IBR.d1') is the one with inverse-distance weights ('listw.d1'). Is is only slightly better than the model with the simple weights ('listw.gab'), whereas the nonspatial model and the one with inverse squared distance weights have much less support. 

Let's have a look at the best model. With the argument 'Nagelkerke = TRUE', we request a pseudo R-squared. 

```{r}
summary(mod.sar.IBR.d1, Nagelkerke = TRUE)
```

- Again, PatchSize is significant but not IBR. 
- The section starting with 'Lamba' summarizes the fitted spatial autocorrelation term. It is not statistically significant (p-value = 0.1039 for the Likelihood Ratio test LR).

## 6. Spatial filtering with MEM using package 'spmoran'

See tutorial for 'spmoran' package: https://arxiv.org/ftp/arxiv/papers/1703/1703.04467.pdf

Both GLS and SAR fitted a spatially correlated error structure of a relatively simple form to the data. Gene flow could be more complex and for example, could create spatial autocorrelation structure that is not the same in all directions or in all parts of the study area. Moran Eigenvector Maps (MEM) allows a more flexible modeling of spatial structure in the data. In spatial filtering, we use MEM spatial eigenvectors to account for any spatial structure while fitting and testing the effect of our predictors.

### Default method

The new package 'spmoran' makes this really easy. First, we create the MEM spatial eigenvectors. This implies defining neighbors and weights, but this is well hidden in the code below. The function 'meigen' here takes the coordinates, calculates a minimum spanning tree (so that each site has at least one neighbour), and finds the maximum distance 'h' from the spanning tree. It then calculates neighbor weights as exp(-dij / h). 

Note: if you have many sites (> 200), the function 'meigen_f' may be used instead of 'meigen', it should even work for >1000 sites.

The function 'esf' then performs the spatial filtering. Here it uses stepwise selection of MEM spatial eigenvectors using an R-squared criterion (fn = "r2").

```{r}
# lm model: using truncated distance matrix (max of min spanning tree distance)
meig <- spmoran::meigen(coords=xy)
e_res <- spmoran::esf( y=Dianthus.df$A, x=Dianthus.df[,c("PatchSize", "IBR")],
                       meig=meig, fn = "r2" )
```

Let's look at some output:

- **e_res$b**: Table with regression results for predictors X.
- **e_res$r**: Table with regression results for MEM spatial eigenvectors.
- **e_res$e**: Summary statistics for the entire model.

```{r}
e_res$b
e_res$r
e_res$e
```

- Again, PatchSize is statistically significant but not IBR.
- 'adjR2' is high (0.437), but this includes the selected MEM spatial eigenvectors! 

### Using a custom connectivity matrix

We know already that 'listw.d1' fit the data well, so let's re-run the model with our own definition of spatial weights. With the funciton 'listw2mat', we convert from 'listw' format to a full connnectivity matrix. 

```{r}
cmat.d1    <- spdep::listw2mat( listw.d1) 
meigW  <- spmoran::meigen( cmat = cmat.d1 )
e_res <- spmoran::esf( y=Dianthus.df$A, x=Dianthus.df[,c("PatchSize", "IBR")],
                       meig=meigW, fn = "r2" )
e_res$b
e_res$r
e_res$e
```

Note: the messages tell us that 'cmat' has been made symmetric before analysis, that 27 out of 59 MEM spatial eigenvector (and their eigenvalues, hence 'pairs') were retained initially and subjected to stepwise selection, which then returned 15 statistically significant MEM eigenvectors that were included in the regression model with the predictor variables X (PatchSize and IBR).

Questions: 

- Does this model fit the data better? Look for a lower AIC. In addition, you can compare the adjusted R-squared. 
- What could cause a difference in model performance?
- Does this affect the results for PatchSize and IBR? Compare both parameter estimates and p-value between two two models.

### Random effect model

The previous model selected 15 MEM spatial eigenvectors, and thus fitted 15 additional models. Just like the random effects for family and population in Week 6 lab, we can save a few parameters here by fitting the set of MEM eigenvectors as a random effect. This is done by the function 'resf'. 

```{r}
r_res <- spmoran::resf( y=Dianthus.df$A, x=Dianthus.df[,c("PatchSize", "IBR")], 
               meig = meigW, method = "reml" ) 
r_res$b
r_res$s
r_res$r
r_res$e
```

Note: we can't compare AIC with the previous models, as the model was fitted with 'reml'. 

We get an additional output 'r_res$s' with two parameters: 

- **shrink_sf_SE**: when squared, this tells us how much variance was captured by the MEM eigenvectors.
- **shrink_sf_alpha**: this give an indication of the spatial scale of the observed pattern. A value above 1 would indicate large-scale spatial structure. Here it seems to be a finer-scale pattern overall.

## 7. Fit spatially varying coefficients model with package 'spmoran'

See: https://arxiv.org/ftp/arxiv/papers/1703/1703.04467.pdf

Now comes the coolest part! 

So far, we have fitted the same model for all sites. Spatially weighted regression would allow relaxing this. Spatial filtering with MEM can be used to accomplish the same goal. The main advantage is that we can visualize how the slope parameter estimates, and their p-values, vary across the study area! This is a great exploratory tool that can help us better understand what is going on. 

### Model with PatchSize and IBR

We fit the model with 'resf_vc'. 

```{r}
rv_res <- spmoran::resf_vc( y=Dianthus.df$A, 
                            x = Dianthus.df[,c("PatchSize", "IBR")], 
                            xconst = NULL, meig = meigW, method = "reml" )
```

Instead of one slope estimate for each predictor, we now get a different estimate for each combination of parameter and site (sounds like overfitting?). Here's a summary of the distbiution of these estimates.

```{r}
summary( rv_res$b_vc ) 
```

The slope estimate for PatchSize varied between -0.0056 and 0.098, with a mean of 0.046. The slope estimate for the 'IBR' term varied between -0.805 and 0.348, with a mean of -0.01! That is an astounding range of variation. Keep in mind that we really expect a positive relationship, there is no biological explanation for a ngeative relationship. 

Here is a similar summary of the p-values:

```{r}
summary( rv_res$p_vc )
```

For both variables, most sites do not show a significant effect (i.e., only few sites show a p-value < 0.05).

We could print these results by site (type 'rv_res$b_vc' or 'rv_res$p_vc'). Even better, we can plot them in space. We start with combining the data ('Dianthus.df') and the results into one data frame 'Results'. By specifying 'b=rv_res$b_vc' and "p=rv_res$p_vc', R will create column names that start with 'b' or 'p', respectively.

```{r}
Result <- data.frame(Dianthus.df, b=rv_res$b_vc, p=rv_res$p_vc)
names(Result)
```

Let's start with PatchSize. Here,  we first plot PatchSize in space, with symbol size as a function of patch size. In a second plot, we color sites by statistical significance and the size of the symbols represents the parameter estimate of the regression slope coefficient for Patch Size. The layer 'coord_fixed' keeps controls the aspect ratio between x- and y-axes. 

```{r fig.show='hold'}
require(ggplot2)
ggplot(as.data.frame(Result), aes(x, y, size=PatchSize)) +
  geom_point(color="darkblue") + coord_fixed()
ggplot(as.data.frame(Result), aes(x, y, col=p.PatchSize < 0.05, size=b.PatchSize)) +
  geom_point() + coord_fixed()
```

Let's do the same for 'IBR':

```{r fig.show='hold'}
require(ggplot2)
ggplot(as.data.frame(Result), aes(x, y, size=IBR)) +
  geom_point(color="darkgreen") + coord_fixed()
ggplot(as.data.frame(Result), aes(x, y, col=p.IBR < 0.05, size=b.IBR)) +
  geom_point() + coord_fixed()
```

- The very small dots in the first map are the ungrazed patches.
- From the second map, it looks like the significant values were the one with negative slope estimates, for which we don't have a biological interpretation.

### Model with IBR only

Keep in mind that 'IBR' and 'PatchSize' showed a strong correlation. The parameter estimates could therefore depend quite a bit on the other variables. To help with the interpretation, let's repeat the last analysis just with 'IBR', without 'PatchSize'. 

```{r}
rv_res <- spmoran::resf_vc( y=Dianthus.df$A, 
                            x = Dianthus.df[,c("IBR")], 
                            xconst = NULL, meig = meigW, method = "reml" )
summary( rv_res$p_vc ) 
```
Now the range of slope estimates is smaller, most sites have a positive estimate.

```{r}
summary( rv_res$p_vc ) 
```

Also, a larger proportion of sites nows has p-values < 0.05.

```{r}
Result <- data.frame(Dianthus.df, b=rv_res$b_vc, p=rv_res$p_vc)
names(Result)

ggplot(as.data.frame(Result), aes(x, y, col=p.V1 < 0.05, size=b.V1)) +
  geom_point() + coord_fixed()
```

This is a very different map of results!

- Most sites now show significant effects. 
- The sites with larger positive estimates show significant effects, whereas those with small or negative estimates show non-significant effects.
- There are 3 - 4 clusters of sites where the IBR models is not effective at explaining variation in allelic richness: in the very East, in the South-East, and one area in the South-West. 
- Knowing the study area, these are distinct regions (e.g. valleys) that may suggest further biological explanations.

## 8. Conclusions

- We moved from pair-wise distance matrices (link-based) to node-based analysis by integrating the explanatory distance matrices for IBD and IBR into patch-level connectivity indices Si (neighborhood analysis). 
- We found no support for the IBD model, and strong support for the IBR model when tested without additional predictors. 
- The site-level predictors 'PatchSize' (log('Ha')) was strongly correlated with our IBR model, and when PatchSize was added to the model, 'IBR' was no long statistically significant and its slope estimate changed considerably.
- The MEM analogue to spatially weighted regression showed very different patterns for 'IBR' depending on whether or not 'PatchSize' was included in the model. Withouth 'PatchSize', 'IBR' showed significant positive correlation with allelic richness across the study area, except for three sub-areas. 
- In practical terms, this may suggest that the management strategy of maintaining plant functional connectivity through shepherding seems to be working for this species overall, though there are three parts of the study area where this may not be sufficient to maintain gene flow. 
- The evidence is not conclusive, however, the observed patterns could also be explained by population size, which in this species seems to be associated with patch size. This makes sense if smaller patches contain smaller populations with higher rates of genetic drift.

```{r message=FALSE, warning=TRUE, include=FALSE}
LandGenCourse::detachAllPackages()
```
